### 2025-05-13 manuscripts/09-JERP-v2/manuscript_main.tex
start working on R&R v2

Understanding discrimination against Hispanics in labor markets has far-reaching implications that extend beyond individual earnings. Labor market discrimination can create persistent intergenerational disadvantages by limiting economic mobility, reducing access to quality healthcare and education, and constraining residential choices \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,bowles2002inheritance, djajic2003assimilation}. Discrimination may foster occupational sorting and segregation, as shown by recent task-based models of racial wage gaps that identify race-specific barriers in “Contact” tasks and explain the stagnation of these gaps post-1980 \autocite{hurst2024task}. These barriers may trap Hispanic families in cycles of lower socioeconomic status, as reduced earnings and employment opportunities limit their ability to invest in their children's human capital or build wealth through homeownership and savings.\footnote{See \textcite{bowles2002inheritance} for an analysis of the intergenerational transmission of economic status.} Although previous research has identified earnings gaps between Hispanic and non-Hispanic Whites, this paper makes several important contributions. First, it develops a novel empirical strategy that better isolates the causal effect of Hispanic ethnicity on educational and labor market outcomes by comparing children from inter-ethnic families. This approach helps control for typically unobservable family background characteristics that might confound traditional analyses. Second, it provides new evidence on the specific role of Hispanic surnames in driving discrimination, offering insights into how ethnic signals influence educational attainment and labor market outcomes. Finally, by examining both education and employment outcomes, this study helps illuminate the channels through which ethnic discrimination may perpetuate economic disparities across generations and extends earlier findings by demonstrating how naming cues can compound structural inequalities and reinforce existing barriers.

This study builds upon and improves the traditional Oaxaca-Blinder-Kitagawa decomposition approach often used in labor market discrimination studies \autocite{kitagawa1955components, oaxaca1973male,blinder1973wage}. While these decomposition methods have provided valuable insights into earnings gaps between groups, including Hispanics and non-Hispanic Whites \autocite{davilaChangesRelativeEarnings2008}, they are limited in their ability to account for unobserved differences between the groups. My approach comparing children of inter-ethnic marriages, provides a more closely matched comparison group, allowing me to better isolate the effect of having a Hispanic last name on labor market outcomes.

@@

Using this novel approach, I find significant effects of Hispanic surnames on both educational and labor market outcomes, and that individuals that likely have a Hispanic last names face substantial disadvantages in educational attainment and earnings compared to their counterparts with likely non-Hispanic names. Specifically, individuals with presumably Hispanic last names complete 0.2 fewer years of education than those with non-Hispanic names, even when controlling for family background. Individuals with Hispanic last names are also 1 percentage point more likely to be unemployed and earn 5 percentage points less than those with non-Hispanic names. These findings suggest that ethnic discrimination continues to play a role in shaping economic opportunities for Hispanics in the United States, highlighting the need for targeted policies to address these persistent disparities.

Factors that affect labor market outcomes, like skills and stereotypes, are unobservable to economists, so it is challenging to identify discrimination. One strategy used by researchers is audit or resume studies. \textcite{bertrand2004emily} conducted an audit study by sending employers identical resumes that differed only in the ethnic and racial signal of the applicant’s name (Black-sounding versus White-sounding). They found that resumes with Black-sounding names received substantially fewer callbacks than their White counterparts. Audit studies only observe callbacks, not wages. The  current study uses a method developed by \textcite{rubinstein2014pride}. The authors compared the children of mixed marriages between Sephardic and Ashkenazi people in Israel.\footnote{Ashkenazi and Sephardic are two distinct Jewish ethnic groups.} They found that workers with Sephardic last names earn substantially less than those with Ashkenazi last names. I compare children of Hispanic fathers and White mothers (henceforth HW) to children of White fathers and Hispanic mothers (WH). This approach accounts for the fact that  couples are likely to have similar income, schooling, and socioeconomic background, so these factors are largely controlled for \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996}.\footnote{For more on assortative mating see \autocite{beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}} Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages (i.e., White fathers-White mothers and Hispanic fathers-Hispanic mothers). In the US, children  from households with a Hispanic father and White mother  are overwhelmingly likely to  have their father’s Hispanic last name, allowing us to investigate how this ethnic signal affects annual log earnings.

@@

I use the CPS for my primary analysis of the effect of a Hispanic surname on earnings. I restrict my sample to US citizens  aged 25–40 born between 1960 and 2000 in the US. Taking advantage of data on parents’ place of birth, I divide the sample into four groups depending on their parents’ ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country or Puerto Rico, and White if they were born in the United States.\footnote{The list of Spanish-speaking countries includes: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item White father and White mother (hereafter WW) 
\item White father and Hispanic mother (hereafter WH)
\item Hispanic father and White mother (hereafter HW)
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

My sample includes inter-ethnic children with Hispanic ancestry aged 25–40 who are US citizens. I include both Hispanic White and non-Hispanic White individuals, since ethnic attrition could lead to biased estimates when measuring gaps as shown  by \textcite{hadah2024effect}. \footnote{A person self-reports Hispanic identity by answering the Census question: ``Is this person Spanish/Hispanic/Latino?''} \footnote{Ethnic attrition happens when a US-born descendant of a Hispanic immigrant fails to self-identify as Hispanic. For more discussion of this phenomenon see \textcite{antmanEthnicAttritionObserved2016,antmanEthnicAttritionAssimilation2020}} To maintain analytical clarity, I restricted the sample to Hispanic and non-Hispanic White individuals, excluding other racial groups who also identify as Hispanic, to avoid confounding racial factors.

@@

Using the “synthetic” parents I constructed using the Census data, I examine the family background of the different types of children. WW couples have the highest level of education among the four groups: 12.58 years  for husbands and 12.36 for wives, so a WW household has 24.95 years of schooling. Men in HH marriages have 8.64 years of education, while women have 8.49. As a household, HH couples have 17.13 years of education. As predicted, interethnic couples marry people similar to them. WH husbands have 11.82 years of education, while wives have an average of 10.71, a household total of 22.68 years. HW husbands have 10.33 years, while wives have an average of 11.01, giving HW households a total of 21.50 years. Both  Hispanic men and women in interethnic couples marry white spouses that are, on average, more educated than members of the average HH couple. More importantly for human capital accumulation, the mother in the HW marriages (those with children most likely to have a Hispanic-sounding last name—are more educated than their WH peers.) Since mothers could be more important in the child’s human capital accumulation and education, this would suggest that a child with a Hispanic last name should complete more years of education  \autocite{gould2020does}.

@@

In terms of labor market performance, WW households boast the highest log total family income at 10.75 , while HH households fall at the lower end with 10.42.  Among interethnic couples, WH households have a slightly higher total income of 10.65 compared to HW households at 10.60. The difference in husbands’ log hourly earnings between HW and WH is marginal, with HW men earning 4\% less than their WH counterparts. HW women surpass WH women in earnings, with HW wives earning 1.75 and WH wives earning 1.73 in log hourly earnings, respectively. This reversal in the typical earning pattern speaks to the closing economic disparities between these groups and implies potentially greater economic contributions from HW women to their families, which could benefit their children when they enter the labor market.

@@

The evidence presented in this section supports a robust empirical strategy, revealing

significant selection in marriage and distinctive educational and income patterns among different types of families. Interethnic couples have comparable levels of education and earnings, suggesting that the children of HW families are likely positioned for better educational outcomes, a critical factor for understanding discrimination in the labor market. The disparities between HW and WH families are considerably less pronounced than those between other groups, emphasizing the importance of comparing these children directly. Such comparisons shed light on the nuanced dynamics of ethnicity, education, and economic outcomes in interethnic marriages.\footnote{I present in Tables \ref{tab:synthmex} and \ref{tab:syntnonmex} the summary statistics which detail the educational and economic profiles of parents from four different ethnic groupings—White White (WW), White Hispanic (WH), Hispanic White (HW), and Hispanic Hispanic (HH) on sub-samples of Hispanics of Mexican and non-Mexican ancestries. I find similar results that describe a selection into inter-ethnic marriages among the two groups.}

@@

Overall, the crude gap between HW and WH workers is equal to 5 percentage points (Table \ref{tab:lastnamereg} column 1). An interethnic worker with a Hispanic last name earns 5 percentage points less than an interethnic worker with a White last name. Even after controlling for hours worked, and including state, year, and age FEs in the estimation, the gap stays at 5 percentage points; however, the difference could be attributed to educational differences. An interethnic individual with a Hispanic last name earns 1 percentage point less than one with a White last name, but the result is statistically insignificant.

>>>>

Understanding discrimination against Hispanics in labor markets has far-reaching implications that extend beyond individual earnings. Labor market discrimination can create persistent intergenerational disadvantages by limiting economic mobility, reducing access to quality healthcare and education, and constraining residential choices \autocite{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,bowles2002inheritance, djajic2003assimilation}. Discrimination may foster occupational sorting and segregation, as shown by recent task-based models of racial wage gaps that identify race-specific barriers in “Contact” tasks and explain the stagnation of these gaps post-1980 \autocite{hurst2024task}. These barriers may trap Hispanic families in cycles of lower socioeconomic status, as reduced earnings and employment opportunities limit their ability to invest in their children's human capital or build wealth through homeownership and savings.\footnote{See \textcite{bowles2002inheritance} for an analysis of the intergenerational transmission of economic status.}

Although previous research has identified earnings gaps between Hispanic and non-Hispanic Whites, this paper makes several important contributions. First, it develops a novel empirical strategy that better isolates the causal effect of Hispanic ethnicity on educational and labor market outcomes by comparing children from inter-ethnic families. This approach helps control for typically unobservable family background characteristics that might confound traditional analyses. Second, it provides new evidence on the specific role of Hispanic surnames in driving discrimination, offering insights into how ethnic signals influence educational attainment and labor market outcomes. Finally, by examining both education and employment outcomes, this study helps illuminate the channels through which ethnic discrimination may perpetuate economic disparities across generations and extends earlier findings by demonstrating how naming cues can compound structural inequalities and reinforce existing barriers.

This study builds upon and improves the traditional Oaxaca-Blinder-Kitagawa decomposition approach often used in labor market discrimination studies \autocite{kitagawa1955components, oaxaca1973male,blinder1973wage}. While these decomposition methods have provided valuable insights into earnings gaps between groups, including Hispanics and non-Hispanic Whites \autocite{davilaChangesRelativeEarnings2008}, they are limited in their ability to account for unobserved differences between the groups. My approach comparing children of inter-ethnic marriages, provides a more closely matched comparison group, allowing me to better isolate the effect of likely having a Hispanic last name on labor market outcomes.

@@

Using this novel approach, I find significant effects of Hispanic surnames on both educational and labor market outcomes, and that individuals that likely have a Hispanic last names face substantial disadvantages in educational attainment and earnings compared to their counterparts with likely non-Hispanic names. Specifically, individuals with presumably Hispanic last names complete 0.2 fewer years of education than those with non-Hispanic names, even when controlling for family background. Individuals with presumably Hispanic last names are also 1 percentage point more likely to be unemployed and earn 5 percentage points less than those with likely non-Hispanic names. These findings suggest that ethnic discrimination continues to play a role in shaping economic opportunities for Hispanics in the United States, highlighting the need for targeted policies to address these persistent disparities.

Factors that affect labor market outcomes, like skills and stereotypes, are unobservable to economists, so it is challenging to identify discrimination. One strategy used by researchers is audit or resume studies. \textcite{bertrand2004emily} conducted an audit study by sending employers identical resumes that differed only in the ethnic and racial signal of the applicant’s name (Black-sounding versus White-sounding). They found that resumes with Black-sounding names received substantially fewer callbacks than their White counterparts. Audit studies only observe callbacks, not wages. The  current study uses a method developed by \textcite{rubinstein2014pride}. The authors compared the children of mixed marriages between Sephardic and Ashkenazi people in Israel.\footnote{Ashkenazi and Sephardic are two distinct Jewish ethnic groups.} They found that workers with Sephardic last names earn substantially less than those with Ashkenazi last names. I compare children of Hispanic fathers and White mothers (henceforth HW) and likely have a Hispanic-sounding last name to children of White fathers and Hispanic mothers (WH). This approach accounts for the fact that  couples are likely to have similar income, schooling, and socioeconomic background, so these factors are largely controlled for \autocite{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996}.\footnote{For more on assortative mating see \autocite{beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}} Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages (i.e., White fathers-White mothers and Hispanic fathers-Hispanic mothers). In the US, children  from households with a Hispanic father and White mother  are overwhelmingly likely to  have their father’s Hispanic last name, allowing us to investigate how this ethnic signal affects annual log earnings.

@@

I use the CPS for my primary analysis of the effect of a Hispanic surname on earnings. I restrict my sample to US citizens  aged 25–40 born between 1960 and 2000 in the US. Taking advantage of data on parents’ place of birth, I divide the sample into four groups depending on their parents’ ethnicity. Mothers or fathers are Hispanic if they were born in a Spanish-speaking country or Puerto Rico, and White if they were born in the United States.\footnote{The list of Spanish-speaking countries includes: Argentina, Bolivia, Chile, Colombia, Costa Rica, Cuba, Dominican Republic, Ecuador, El Salvador, Guatemala, Equatorial Guinea, Honduras, Mexico, Nicaragua, Panama, Paraguay, Peru, Spain, Uruguay, Venezuela.} Therefore, an observation can be the product of four types of parents:
\begin{enumerate}
\item White father and White mother (hereafter WW) 
\item White father and Hispanic mother (hereafter WH)---the comparison group and the group that is likely to have a non-Hispanic-sounding last name
\item Hispanic father and White mother (hereafter HW)---the focus of this study and the group most likely to have a Hispanic-sounding last name
\item Hispanic father and Hispanic mother (hereafter HH).
\end{enumerate}

My sample includes inter-ethnic children with Hispanic ancestry aged 25–40 who are US citizens. I include both Hispanic White and non-Hispanic White individuals in this study. For second-generation immigrants, I use parents' place of birth as a proxy for ethnicity. This approach helps avoid biased estimates that could result from ethnic attrition, as demonstrated by \textcite{hadah2024effect}. \footnote{A person self-reports Hispanic identity by answering the Census question: ``Is this person Spanish/Hispanic/Latino?''} \footnote{Ethnic attrition happens when a US-born descendant of a Hispanic immigrant fails to self-identify as Hispanic. For more discussion of this phenomenon see \textcite{antmanEthnicAttritionObserved2016,antmanEthnicAttritionAssimilation2020}} To maintain analytical clarity, I restricted the sample to Hispanic and non-Hispanic White individuals, excluding other racial groups who also identify as Hispanic, to avoid confounding racial factors.

@@

Using the “synthetic” parents I constructed using the Census data, I examine the family background of the different types of children. WW couples have the highest level of education among the four groups: 12.58 years  for husbands and 12.36 for wives, so a WW household has 24.95 years of schooling. Men in HH marriages have 8.91 years of education, while women have 8.68. As a household, HH couples have 17.69 years of education. As predicted, interethnic couples marry people similar to them. WH husbands have 11.82 years of education, while wives have an average of 10.71, a household total of 22.68 years. HW husbands have 10.33 years, while wives have an average of 11.01, giving HW households a total of 21.50 years. Both  Hispanic men and women in interethnic couples marry white spouses that are, on average, more educated than members of the average HH couple. More importantly for human capital accumulation, the mother in the HW marriages (those with children most likely to have a Hispanic-sounding last name—are more educated than their WH peers.) Since mothers could be more important in the child’s human capital accumulation and education, this would suggest that a child with a Hispanic last name should complete more years of education  \autocite{gould2020does}.

@@

In terms of labor market performance, WW households boast the highest log total family income, while HH households fall at the lower end.  Among interethnic couples, WH households have a slightly higher total income compared to HW households---WH households earn 5\% more than HW. The difference in husbands’ log hourly earnings between HW and WH is marginal, with HW men earning 4\% less than their WH counterparts. HW women earn 2\% more than WH women. This reversal in the typical earning pattern speaks to the closing economic disparities between these groups and implies potentially greater economic contributions from HW women to their families, which could benefit their children when they enter the labor market.

@@

The evidence presented in this section supports a robust empirical strategy, revealing significant selection in marriage and distinctive educational and income patterns among different types of families. Interethnic couples have comparable levels of education and earnings, suggesting that the children of HW families are likely positioned for better educational outcomes, a critical factor for understanding discrimination in the labor market. The disparities between HW and WH families are considerably less pronounced than those between other groups, emphasizing the importance of comparing these children directly. Such comparisons shed light on the nuanced dynamics of ethnicity, education, and economic outcomes in interethnic marriages.\footnote{I present in Tables \ref{tab:synthmex} and \ref{tab:syntnonmex} the summary statistics which detail the educational and economic profiles of parents from four different ethnic groupings—White White (WW), White Hispanic (WH), Hispanic White (HW), and Hispanic Hispanic (HH) on sub-samples of Hispanics of Mexican and non-Mexican ancestries. I find similar results that describe a selection into inter-ethnic marriages among the two groups.}

@@

Overall, the crude gap between HW and WH workers is equal to 5 percentage points (Table \ref{tab:lastnamereg} column 1). An interethnic worker with a Hispanic last name earns 5 percentage points less than an interethnic worker with a White last name. Even after controlling for hours worked, and including state, year, and age FEs in the estimation, the gap stays at 5 percentage points; however, the difference could be attributed to educational differences. An interethnic man with a likely Hispanic last name earns 1 percentage point less than one with a White last name, but the result is statistically insignificant.
